---
title: "Getting Started with dagriculture"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with dagriculture}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

`dagriculture` is a pure, value-oriented graph library. It is designed to manage graph structures, topological dependencies, structural validity, and task planning. It does not natively orchestrate runtime execution or manage side effects; rather, it tracks structural states (`new`, `ready`, `blocked`) and enforces declarative dependency rules.

## Core Concepts

*   **Registry:** Defines the allowable `kinds` of nodes in a graph.
*   **Graph:** The main immutable data structure carrying nodes, edges, and gates. Operations return a new graph and bump the internal `$version`.
*   **Nodes & Edges:** Represent units of computation (or data) and their dependencies.
*   **Gates:** Explicit structural blockers attached to edges, which prevent downstream nodes from becoming `ready` until the gate is resolved.
*   **Planning:** Topologically sorting and tracking the structural readiness of nodes.
*   **Diffing:** Comparing two graph values id by id, down to individual field changes.
*   **Rendering:** Turning a graph value into text formats such as Mermaid flowcharts.

## Installation

```r
pak::pak("sims1253/dagriculture")
```

## Basic Workflow

### 1. Define a Registry

First, establish a registry to strictly define what kind of nodes are allowed in your graphs.

```{r}
library(dagriculture)

reg <- dagri_registry(
  dagri_kind("input", output_type = "data.frame"),
  dagri_kind("model"),
  dagri_kind("report")
)
```

### 2. Build the Graph

Start with an empty graph, and add nodes and edges. Because `dagriculture` uses value semantics, every mutating function returns a new updated graph instance.

```{r}
g <- dagri_graph(reg)

g <- g |>
  dagri_add_node(id = "data", kind = "input", label = "Raw Data") |>
  dagri_add_node(id = "fit", kind = "model", label = "Statistical Fit") |>
  dagri_add_node(id = "plot", kind = "report", label = "Final Plot") |>
  dagri_add_edge(from = "data", to = "fit", id = "e_data_fit") |>
  dagri_add_edge(from = "fit", to = "plot", id = "e_fit_plot")

# The internal version tracks changes
g$version
```

### 3. Add a Gate

You can place a "gate" on an edge to represent a manual or external block—for instance, requiring human approval before the statistical fit proceeds.

```{r}
g <- dagri_add_gate(g, edge_id = "e_data_fit", id = "approval_gate")
```

## Structural State Semantics

`dagriculture` tracks **structural** readiness only. It does not execute anything and
does not know whether a node has been computed, is currently running, or has failed.
The three structural node states, set by [dagri_recompute_state()], are:

| `state`    | Meaning                                                                                  |
|------------|------------------------------------------------------------------------------------------|
| `new`      | Initial state of a freshly added node; `dagri_recompute_state()` overwrites it on the next pass. |
| `ready`    | Every upstream node is `ready` and no inbound edge carries a pending gate; the node is structurally eligible to run. |
| `blocked`  | At least one structural blocker prevents readiness. See `block_reason` below.            |

When a node is `blocked`, `dagri_recompute_state()` records a `block_reason`:

| `block_reason`       | Meaning                                                                                  |
|----------------------|------------------------------------------------------------------------------------------|
| `none`              | Not blocked (set together with `state = "ready"`).                                        |
| `gate`              | An inbound edge carries one or more gates with `status = "pending"`.                      |
| `upstream_blocked`  | At least one upstream node is itself `blocked`, so this node cannot become `ready` yet.   |

### What `dagri_recompute_state()` does NOT track

Execution state is the consumer's overlay, not part of the structural model. In
particular, `dagriculture` never records:

| Execution concept (consumer-owned) | Why it is out of scope                                                     |
|------------------------------------|---------------------------------------------------------------------------|
| `done` / `succeeded`               | A result existing is an artifact/cache concern (bayesgrove overlays this). |
| `failed`                           | Failure is a runtime outcome; the graph only knows structural readiness.   |
| `running`                          | Concurrency and scheduling live in the executor, not the graph.            |
| `skipped` (cache hit)              | Reuse is decided from fingerprints, not from node `$state`.               |

A node that is structurally `ready` may still be skipped by a consumer when a
reusable result exists; conversely a `blocked` node will never run until its
blocker clears. Call [dagri_recompute_state()] whenever the graph structure or
gate statuses change, so the `state`/`block_reason` fields reflect the current
graph before planning.

### 4. Compute State and Plan

Use `dagri_recompute_state()` to evaluate the graph topologically and determine which nodes are structurally ready and which are blocked.

```{r}
g_state <- dagri_recompute_state(g)

# The 'data' node is a root and has no blockers
dagri_eligible(g_state)

# The 'fit' node is blocked by the gate; the 'plot' node is blocked because 'fit' is blocked.
dagri_blocked(g_state)
```

We can generate a structural plan to see how dependencies shake out. `dagri_plan()`
derives state internally, so it does not require the recomputed graph value
above:

```{r}
plan <- dagri_plan(g_state)
plan$targets
plan$blocked
plan$external_blocked
plan$pending_gates
```

Planner-visible external holds can be layered onto planning without mutating the
graph:

```{r}
dagri_plan(
  g_state,
  external_holds = list(fit = "manual_review")
)$external_blocked
```

### 5. Resolve the Gate

Once the external condition is met, you can resolve the gate. Computing the state again reveals that the downstream dependencies are now unblocked.

```{r}
g_unblocked <- dagri_resolve_gate(g_state, id = "approval_gate") |>
  dagri_recompute_state()

# Everything is unblocked and ready for execution
dagri_eligible(g_unblocked)
dagri_blocked(g_unblocked)
```

## Querying Topology

`dagriculture` includes standard helpers for topological traversal and queries.

```{r}
dagri_upstream(g_unblocked, "plot")
dagri_descendants(g_unblocked, "data")
dagri_leaves(g_unblocked)
dagri_topo_order(g_unblocked)
```

## Branching and Comparing Graphs

Because graphs are immutable values, exploring an alternative is cheap: bind the
current graph to a new name and mutate from there. Every mutator returns a fresh
graph value, so the shared history is never rewritten and the branches can be
kept side by side.

```{r}
# Both branches start from `g`, where the approval gate is still pending.

# Branch A: the reviewer signs off and the scope grows a summary node.
g_signed <- g |>
  dagri_resolve_gate(id = "approval_gate") |>
  dagri_add_node(id = "summary", kind = "report", label = "Executive Summary") |>
  dagri_add_edge(from = "fit", to = "summary", id = "e_fit_summary")

# Branch B: a quick-look variant that plots straight from the raw data.
g_quick <- g |>
  dagri_remove_edge(id = "e_fit_plot") |>
  dagri_add_edge(from = "data", to = "plot", id = "e_data_plot") |>
  dagri_update_node(id = "plot", label = "Quick Look")
```

[dagri_graph_diff()] reports what it takes to go from one graph value
(`before`) to another (`after`). The result is a pure structural comparison —
nine id vectors, no planning or workflow semantics attached:

```{r}
diff <- dagri_graph_diff(g_quick, g_signed)

diff$added_nodes    # "summary" exists only in g_signed
diff$removed_nodes  # empty: neither branch drops a node
diff$added_edges    # "e_fit_plot" and "e_fit_summary" exist only in g_signed
diff$removed_edges  # "e_data_plot" exists only in g_quick
diff$added_gates    # empty: the gate exists in both branches
diff$removed_gates  # empty
```

Read the direction as `before -> after`: `added_*` ids are present only in
`after`, `removed_*` ids only in `before`. Ids present in both graphs can still
differ field by field, which the `changed_*` vectors report:

```{r}
diff$changed_nodes  # "plot": its label differs between the branches
diff$changed_edges  # empty
diff$changed_gates  # "approval_gate": pending in g_quick, resolved in g_signed
```

The tracked fields are the node
`kind`/`label`/`params`/`state`/`block_reason`/`metadata`, the edge
`from`/`to`/`type`/`metadata`, and the gate `edge_id`/`status`/`metadata`. The
internal `$version` is deliberately ignored — it is derived bookkeeping, not
content. Because node `state` and `block_reason` are tracked, recomputing state
on one branch only (as [dagri_recompute_state()] did above) also surfaces as
`changed_nodes`; that is still structural bookkeeping, not an execution record.

To see *what* changed, not just *where*, pass `include_values = TRUE`. The
result additionally carries a `values` element: a named list with `nodes`,
`edges`, and `gates`, each keyed by the ids of the matching `changed_*` vector
(same ids, same order). Every entry lists only the differing fields, each as
`list(before = ..., after = ...)`:

```{r}
value_diff <- dagri_graph_diff(g_quick, g_signed, include_values = TRUE)

# The plot node's label on the way from g_quick to g_signed:
value_diff$values$nodes$plot$label

# The gate flipped from pending to resolved:
value_diff$values$gates$approval_gate$status
```

Ids that were only added or removed do not appear in `values` — there is
nothing to compare them against. Edge fields follow the same shape; for
example, re-typing an edge shows up like this:

```{r}
g_control <- dagri_update_edge(g, "e_fit_plot", type = "control")
edge_diff <- dagri_graph_diff(g, g_control, include_values = TRUE)

edge_diff$changed_edges
edge_diff$values$edges$e_fit_plot
```

## Visualizing with Mermaid

[dagri_mermaid()] renders any graph value as Mermaid flowchart text: one line
per node carrying its label (the id when no label is set), a companion `class`
line named after the node's structural `state`, and one arrow per edge. Pending
gates annotate the edge they block:

```{r}
cat(dagri_mermaid(g_state))
```

The `class` lines are inert until a `classDef` with a matching name exists,
which is what the `class_defs` argument is for. Gate annotation text defaults
to `gate: <id>` but can be replaced with the `gate_label` callback, which
receives the full gate and edge records — so it can map `status` to review
language or pull a reviewer name from `gate$metadata`:

```{r}
cat(dagri_mermaid(
  g_state,
  gate_label = function(gate, edge) {
    if (identical(gate$status, "resolved")) "signed off" else "pending review"
  },
  class_defs = c(
    "classDef ready fill:#e8f5e9",
    "classDef blocked fill:#ffebee",
    "classDef new fill:#fff3e0"
  )
))
```

Resolved gates are omitted by default — they no longer block anything. With
`include_resolved_gates = TRUE` they rejoin the annotation, each suffixed with
a renderer-owned `" (resolved)"` marker:

```{r mermaid-embed, echo = FALSE, comment = NA, collapse = FALSE, class.output = "mermaid"}
cat(dagri_mermaid(
  g_unblocked,
  gate_label = function(gate, edge) "signed off",
  include_resolved_gates = TRUE
))
```

The chunk above sets `class.output = "mermaid"`, so its output block carries
the `mermaid` class: documentation sites that load mermaid.js turn such blocks
into live diagrams. pkgdown's stock Bootstrap 5 template does not ship
mermaid.js, so on a default pkgdown site — and in plain R documentation — the
block simply stays the readable flowchart text you see above; a site owner can
opt in to live rendering by adding mermaid.js to the site head (for example via
pkgdown's `includes`).

Two properties worth knowing when embedding the output. Node labels, gate
annotations, and `class_defs` lines are sanitized (`"` becomes `'`; brackets,
braces, pipes, and angle brackets become spaces), so prefer paren-free color forms
such as `fill:#3480db66` over `rgba(...)`, whose parentheses would be stripped.
And emission follows graph insertion order deterministically, so the same graph
value always renders to the same string — snapshot-friendly. Node ids are
emitted verbatim as Mermaid identifiers, so keep them alphanumeric (the ids in
this vignette are safe by construction).
