---
title: "Serialization and the Consumer Boundary"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Serialization and the Consumer Boundary}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(dagriculture)
library(jsonlite)
```

`dagriculture` graphs are plain named lists, so they serialize with any
JSON writer. In practice, though, every consumer that persists graphs with
jsonlite rediscovers the same handful of rules: which fields are required,
how `NULL` round-trips, why map keys and embedded ids both exist, and what
the package's validation actually checks when a graph comes back on load.
This vignette records those rules once, as a canonical writer and reader
recipe you can copy.

The design backdrop is `design/persistence-spec.md` in the repository;
nothing here is new policy — it is that policy applied to jsonlite.

## The graph snapshot shape

A graph is a named list with five fields the entry-point validation
requires (`registry`, `nodes`, `edges`, `gates`, `version`) plus the graph-level
`metadata` that every record type carries. Starred fields are nullable —
they hold R `NULL` when absent:

- **graph**: `registry`, `nodes`, `edges`, `gates`, `version`, `metadata`
- **registry**: `kinds`, `metadata`
- **kind**: `name`, `input_contract`*, `output_type`*, `param_schema`*,
  `metadata`
- **node**: `id`, `kind`, `label`*, `params`, `state`, `block_reason`,
  `metadata`
- **edge**: `id`, `from`, `to`, `type`, `metadata`
- **gate**: `id`, `edge_id`, `status`, `metadata`

Three shape details matter for serialization:

- **`registry`, `nodes`, `edges`, and `gates` are named maps keyed by id**
  (kinds by kind name), not arrays of records. Empty maps are *named, empty
  lists* — `setNames(list(), character(0))` — and serialize as `{}`, not `[]`.
- **Nullable fields hold R `NULL`** when absent, and must come back as R
  `NULL` after a round trip, not as empty lists.
- **`metadata` is opaque caller-owned plain data** on every record: a named
  list whose values contain no closures, environments, formulas, language
  objects, S4 objects, external pointers, or weak references, at any depth.

`state`/`block_reason`/`status` are plain strings with fixed enums
(`new`/`ready`/`blocked`, `none`/`gate`/`upstream_blocked`,
`pending`/`resolved`); `version` is a single R integer that counts edits.

```{r snapshot-example}
reg <- dagri_registry(
  dagri_kind("source", output_type = "data.frame"),
  dagri_kind("fit", param_schema = list(required = c("model")))
)

g <- dagri_graph(reg, metadata = list(project = "pilot")) |>
  dagri_add_node("raw", "source", label = "Raw Data") |>
  dagri_add_node("m1", "fit", params = list(model = "baseline")) |>
  dagri_add_edge("raw", "m1", id = "e1") |>
  dagri_add_gate("e1", id = "g1")

names(g)
names(g$nodes)
g$nodes$m1$label           # contractually nullable -> NULL
g$registry$kinds$source$param_schema  # also nullable -> NULL
```

Note that `dagriculture` itself does not add `schema_name`/`schema_version`
wrapper fields. If you persist snapshots under the persistence spec, wrap the
graph yourself (for example `schema_name = "dagri_graph_snapshot"`,
`schema_version = 1`) and unwrap before hydrating.

## Writing: one canonical call

```{r writer}
write_graph_json <- function(graph, file = NULL, pretty = FALSE) {
  txt <- toJSON(
    graph,
    auto_unbox = TRUE,
    null = "null",
    digits = NA,
    pretty = pretty
  )
  if (is.null(file)) {
    txt
  } else {
    writeLines(txt, file)
    invisible(file)
  }
}
```

Each option is load-bearing; the jsonlite defaults break round trips:

- `auto_unbox = TRUE` writes length-1 atomic vectors as JSON scalars
  (`"state": "new"` instead of `["new"]`).
- `null = "null"` writes R `NULL` as JSON `null`. The jsonlite *default*
  (`null = "list"`) emits `{}`, which reads back as an empty list, not
  `NULL` — the classic broken round trip.
- `digits = NA` writes doubles at full precision. The default `digits = 4`
  silently truncates `1.23456789` to `1.2346`.

```{r writer-demo}
txt <- write_graph_json(g)
cat(txt)
```

Two caveats to know about, both consequences of JSON being shape-blind:

- `auto_unbox = TRUE` cannot distinguish a scalar from a length-1 vector:
  `params = list(cols = c("x"))` writes `"cols": "x"`. If a length-1 array
  must stay an array, wrap it in `I()` (`params = list(cols = I(c("x")))`
  writes `"cols": ["x"]`).
- An empty *unnamed* list writes as `[]`, an empty *named* list as `{}`.
  dagriculture's collections are always named, so they emit `{}`. The
  default `metadata = list()` is unnamed-empty and emits `[]`; both forms
  read back as empty lists and both are valid metadata.

For reproducible bytes (hashing, diffing), pin every option — including
`pretty` — and sort your records before serializing. `toJSON()` preserves
the insertion order of the R list; the persistence spec asks writers to emit
keys in stable lexical order, so call `graph$nodes <- graph$nodes[sort(names(graph$nodes))]`
(and likewise for edges, gates, kinds) when byte stability matters.

## Reading: simplify, normalize, check, validate

The reader is the mirror image, plus three fixes jsonlite will not do for
you. Read with `simplifyVector = FALSE`, normalize JSON `null` back to the
nullable R fields, check map keys against embedded ids, then confirm the
revived graph through a public call — the recipe ends with
`dagri_topo_order()`, which exercises the validation every public function
runs on entry and additionally rejects cyclic snapshots.

`fromJSON(txt, simplifyVector = FALSE)` keeps every JSON object a named list
and never coerces arrays or heterogeneous values. With the default
simplification, `["a", null, "b"]` collapses to `c("a", NA, "b")` — your
`NULL` param value silently becomes `NA` — and length-1 arrays collapse to
scalars.

```{r reader-helpers}
# JSON null (and the {} that non-canonical writers emit for absent nullable
# fields) must land as R NULL on the contractually nullable fields.
normalize_nulls <- function(graph) {
  nullable_kind_fields <- c("input_contract", "output_type", "param_schema")
  for (i in seq_along(graph$registry$kinds)) {
    kind <- graph$registry$kinds[[i]]
    for (field in nullable_kind_fields) {
      if (is.list(kind[[field]]) && length(kind[[field]]) == 0L) {
        kind[[field]] <- NULL
      }
    }
    graph$registry$kinds[[i]] <- kind
  }
  for (i in seq_along(graph$nodes)) {
    node <- graph$nodes[[i]]
    if (is.list(node$label) && length(node$label) == 0L) {
      node$label <- NULL
    }
    graph$nodes[[i]] <- node
  }
  graph
}

# Named-map keys are the canonical lookup; the embedded id/name fields must
# agree with them (see the next section).
check_snapshot_maps <- function(graph) {
  stopifnot(
    identical(names(graph$nodes), unname(vapply(graph$nodes, `[[`, character(1), "id"))),
    identical(names(graph$edges), unname(vapply(graph$edges, `[[`, character(1), "id"))),
    identical(names(graph$gates), unname(vapply(graph$gates, `[[`, character(1), "id"))),
    identical(
      names(graph$registry$kinds),
      unname(vapply(graph$registry$kinds, `[[`, character(1), "name"))
    )
  )
  invisible(TRUE)
}

read_graph_json <- function(txt) {
  graph <- fromJSON(txt, simplifyVector = FALSE)
  graph <- normalize_nulls(graph)
  check_snapshot_maps(graph)
  # Restore the S3 class so print.dagri_graph() dispatches.
  structure(graph, class = c("dagri_graph", "list"))
}
```

```{r reader-demo}
g2 <- read_graph_json(txt)

is.null(g2$nodes$m1$label)                  # nullable field restored as NULL
is.null(g2$registry$kinds$source$param_schema)
identical(names(g2$gates), "g1")

dagri_topo_order(g2)
```

`dagri_topo_order()` is the public load check: every `dagriculture` function
validates graph structure on entry, and topological ordering additionally
rejects cyclic snapshots — exactly the hydration hazard a revived file can
carry. A revived graph that sorts cleanly is structurally sound *and* acyclic.
The validation is deliberately kept cheap and structural, so note what it
does *not* do (next sections).

One asymmetry to remember when comparing: `toJSON()` writes R atomic vectors
as JSON arrays, and `simplifyVector = FALSE` reads JSON arrays back as R
*lists*. So `param_schema = list(required = c("model"))` round-trips as
`list(required = list("model"))` — semantically equal, not `identical()`.
Normalize both sides (for example, re-serialize each with
`write_graph_json()` and compare the text) instead of calling `identical()`
on a reloaded graph.

## Named-map keys and embedded ids

Both exist on purpose. The map keys (`names(graph$nodes)`) are what
dagriculture itself looks things up by — `dagri_node()`, referential checks
in the entry-point validation, duplicate detection in the editors. The
embedded `id` (or `name`, for kinds) travels *inside* the record so
snapshot consumers that iterate records without a map context (JSONL
exports, database rows) still know what they are looking at.

The entry-point validation uses the map keys for referential integrity but
does not compare them to the embedded ids, so a hand-edited file where the
two disagree passes every public call and misbehaves later. That is what
`check_snapshot_maps()` above guards:

```{r key-check-demo}
check_snapshot_maps(g2)

# A key/id mismatch is invisible to the entry-point validation: it resolves
# references by map key and never reads the embedded id. Rename the key of
# an edge-less node and public calls still pass...
tampered <- dagri_add_node(g2, "iso", "source")
names(tampered$nodes)[[3]] <- "renamed"
dagri_topo_order(tampered)
```

```{r key-check-error, error = TRUE}
# ...but the map check catches it:
check_snapshot_maps(tampered)
```

## Choosing ids

The *enforced* rule is narrow: every id-taking argument of the constructors
and editors (`dagri_kind()`'s `name`, `dagri_add_node()`'s `id`,
`dagri_add_edge()`'s `from`/`to`/`id`, `dagri_add_gate()`'s `edge_id`/`id`,
and the update/remove functions) must be a single non-empty, non-`NA`
character string. dagriculture does not restrict the character set and does
not sanitize ids on load.

The *recommended* pattern is `[A-Za-z0-9_.-]+`. Ids become JSON object keys,
file names, and (via `dagri_mermaid()`) Mermaid node identifiers, which
dagriculture does not sanitize — ids outside that pattern work in memory but
tend to break one of those consumers. Two more key-shape rules worth
respecting:

- Keep map keys unique. R named lists can technically carry duplicate names;
  JSON objects cannot, and duplicate keys silently collapse to the last
  value on read. The editors already reject duplicate node/edge/gate ids at
  edit time.
- Remember that ids are matched exactly, never normalized: `"fit_1"` and
  `"fit.1"` are different nodes.

## The no-code boundary on load

The persistence rule is: persist plain data, never executable code. JSON
gives you that by construction — a closure cannot survive `toJSON()`. An
RDS file can, which is where the boundary needs stating precisely.

Every public `dagriculture` function runs the same cheap structural check
before doing anything with a graph — the validator behind it
(`dagri_validate_graph()`) is internal and unexported, so public calls are
the boundary. It checks structure only:

1. required top-level fields are present and components are lists (not
   data.frames),
2. `version` is a single non-`NA` integer,
3. every edge's `from`/`to` resolves against `names(graph$nodes)`,
4. every gate's `edge_id` resolves against `names(graph$edges)`.

It does **not** scan `params`, `metadata`, or `param_schema` for closures or
other reference-bearing values, and it does not detect cycles (that is
`dagri_topo_order()`'s job when you traverse). Concretely, a graph smuggled
through `saveRDS()` with a closure in `params` passes every structural
check:

```{r rds-demo}
suspicious <- dagri_graph(reg) |>
  dagri_add_node("raw", "source", params = list(loader = function(x) x))

f <- tempfile(fileext = ".rds")
saveRDS(suspicious, f)
loaded <- readRDS(f)

class(loaded$nodes$raw$params$loader)  # the closure survived the round trip
dagri_topo_order(loaded)               # ...and the public surface accepts it
```

So where does the no-code guarantee actually come from? From the *entry
points*, not from the load boundary:

- `metadata` is validated as named plain data by every constructor and
  editor that accepts it (`dagri_kind()`, `dagri_registry()`, `dagri_graph()`,
  `dagri_add_node()`, `dagri_update_node()`, `dagri_add_edge()`,
  `dagri_update_edge()`, `dagri_add_gate()`, `dagri_update_gate()`):
  closures, environments, formulas, language objects, S4 objects, external
  pointers, and weak references are rejected recursively.
- `param_schema` is closure-checked by `dagri_kind()`.

Values that never passed through an entry point — an RDS payload, or direct
assignment like `graph$nodes$raw$metadata <- list(env = new.env())` — are
not re-scanned on load or on the next public call. Practical guidance:
treat the JSON recipe above as your persistence format and RDS only as a
trusted-cache format. If you must load graphs from RDS you do not fully
trust, re-check metadata yourself before use (the same recursive plain-data
test the entry points apply), or push each record's caller-owned fields back
through a public mutator, which validates what it accepts.

## `param_schema` is declarative

A kind's `param_schema` is documentary. `dagri_kind()` validates its
*shape* — a named list, arbitrarily nested, free of closures — but
dagriculture does **not** enforce schema constraints against node `params`,
ever. Nothing stops `dagri_add_node(g, "m1", "fit", params = list(wrong = 1))`
under a kind whose `param_schema` says `required = c("model")`.

The one params check that does exist is a different, weaker mechanism:
`dagri_add_node()` enforces the kind's `input_contract` — params must
*contain* the contract's field names (types are not checked). If you need
`param_schema` actually enforced, validate node params against it in your
own constructor or wrapper before they enter a graph.

## Comparing and hashing after a reload

- Fix the writer options (`auto_unbox`, `null`, `digits`, `pretty`) — any
  change rewrites every serialized value.
- Normalize before comparing: run the reader recipe on both sides so
  nullable fields are `NULL`, keys are checked, and vectors-versus-lists
  differences are resolved.
- `graph$version` is a reliable structural edit counter, but it is not a
  content hash: two graphs built in different orders can differ in
  `version`, map order, and `state` while being structurally equal.
